Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-05 sources/bmc_template_submission/bmc_article.tex
math formatting

%%% Load packages
\usepackage{amsthm,amsmath}
%\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

\usepackage{hyperref}

@@

The most recent *Ilex* tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{study ot\_1984}
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
<!-- **MTH: you probably need some details about the hardware, given the fact
hat you are discussing running times** -->
The updated \cite{gottlieb2005molecular} tree (Fig. \@ref(fig:results)) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional *Ilex* taxa. The best RaxML tree is 99% resolved, with 25% of nodes
with bootstrap support < 0.1 and 48% nodes with bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches
< 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75
(Figure 2).
For comparison, \cite{yao2020phylogeny} also contains all 41 distinct taxa from
the original
\cite{gottlieb2005molecular} study,
and contributes phylogenetic data to 134 additional *Ilex* taxa, from which
`r 135-68` are also in updated \cite{gottlieb2005molecular}. While \cite{yao2020phylogeny} also used
ITS as a marker, their GenBank data is not released yet, so Physcraper was unable
to incorporate 68 additional taxa into the analysis. However, Physcraper was able
to incorporate 18 taxa that were not in \cite{yao2020phylogeny}.

\section*{Discussion}
Text for this section\ldots

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
%\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

%\usepackage{hyperref}

@@

The most recent \textit{Ilex} tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{study ot\_1984}
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact hat you are discussing running times -->
The updated \cite{gottlieb2005molecular} tree (Figure 2) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best RaxML tree is 99\% resolved, with 25\% of nodes
with bootstrap support $<$ 0.1 and 48\% nodes with bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
< 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
For comparison, \cite{yao2020phylogeny} also contains all 41 distinct taxa from
the original
\cite{gottlieb2005molecular} study,
and contributes phylogenetic data to 134 additional \textit{Ilex} taxa, from which
`r 135-68` are also in updated \cite{gottlieb2005molecular}. While \cite{yao2020phylogeny} also used
ITS as a marker, their GenBank data is not released yet, so Physcraper was unable
to incorporate 68 additional taxa into the analysis. However, Physcraper was able
to incorporate 18 taxa that were not in \cite{yao2020phylogeny}.

\section*{Discussion}

Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge are needed.

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of
public molecular data (e.g., Phylota \cite{sanderson2008phylota},
PHLAWD \cite{smith2009mega},
SUPERSMART \cite{antonelli2017toward}) focus on generating full phylogenies \textit{de novo},
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates.
While Physcraper does not generate phylogenies \textit{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.
The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

Physcraper generates individual gene trees, failing to capture the complexity of
species' evolutionary history \cite{song2012resolving}. Yet, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that can be used to reconstruct species trees with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).

Physcraper can potentially link phylogenies to data available in any of the
taxonomies integrated in the OpenTree taxonomy \cite{rees2017automated}, such as
geographical locations from the Global Biodiversity Information Facility, or
fossils from the Paleobiology Database.
%% a resource that is contributing to achieve the goal of "exploring and
%% analysing biodiversity at an accelerated pace, and returning systematics into
%% the mainstream of science" \cite{wilson2003encyclopedia}.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
systematizing molecular (and other) databases, i.e., curating taxonomic assignations
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.
